--- 
title: "Phylogenetic Biology"
author: "Casey W. Dunn"
github-repo: caseywdunn/phylogenetic_biology
twitter-handle: caseywdunn
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "An introduction to Phylogenetic Biology."
fig_caption: yes
---

```{r preliminaries, include=FALSE}
	# The following should be installed from github with the specified 
	# devtools command. You will need to install devtools first.
	library( treeio ) # devtools::install_github("GuangchuangYu/treeio")
	library( ggtree ) # devtools::install_github("GuangchuangYu/ggtree")
	library( ape )
  library( geiger )
	library( gridExtra )
	library( magrittr )
	library( phytools )
	library( stringr )
	library( tidyverse )
	source( "functions.R" )
	set.seed( 23456 )
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  dpi = 300,
  cache = FALSE
  )
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'ape', 'geiger', 'ggtree', 'treeio', 'phytools'
), 'packages.bib')
```


# Preface {-}

I developed this book as a collection of the concepts, methods, and applications that I most wanted to share with students in my Phylogenetic Biology courses. It is really my lecture notes dessed up in the form of the rough draft of a book.

## Source code

The source code for this manuscript is available on [github]( https://github.com/caseywdunn/phylogenetic_biology). The book was rendered from the source with `bookdown` [@bookdown2016].

Please submit any errors you find, typos, or suggestions that you have for improving the manusctipt to the [issue tracker](https://github.com/caseywdunn/phylogenetic_biology/issues).


## Software versions

This book was rendered from the source code on `r format( Sys.time(), "%a %b %d %X %Y" )` with the following R package versions.

```{r session_summary, echo=FALSE, comment=NA}
	sessionInfo()
```


## License

This manuscript is distrubuted under a [Creative Commons Attribution-NonCommercial-NoDerivs 3.0 License]( https://creativecommons.org/licenses/by-nc-nd/3.0/us/).
