--- 
title: "Phylogenetic Biology"
author: "Casey W. Dunn"
github-repo: caseywdunn/phylogenetic_biology
twitter-handle: caseywdunn
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "An introduction to Phylogenetic Biology."
fig_caption: yes
---

```{r preliminaries, include=FALSE}
	# The following should be installed from github with the specified 
	# devtools command. You will need to install devtools first.
	library( treeio ) # devtools::install_github("GuangchuangYu/treeio")
	library( ggtree ) # devtools::install_github("GuangchuangYu/ggtree")
	library( ape )
  library( geiger )
	library( gridExtra )
	library( magrittr )
	library( phytools )
	library( stringr )
	library( tidyverse )
  library( Matrix )
  library( phangorn )
  library( kableExtra )
  library( ggrepel )

	source( "functions.R" )
	set.seed( 23456 )
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  dpi = 300,
  cache = FALSE
  )
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'ape', 'geiger', 'ggtree', 'treeio', 'phytools'
), 'packages.bib')
```


# Preface {-}

I developed this book as a collection of the concepts, methods, and applications that I most wanted to share with students in my Phylogenetic Biology courses. It is really my lecture notes dressed up in the form of the rough draft of a book.

## Additional resources

Please see the appendices for a variety of additional resources.

## Acknowledgements

Thanks in particular to the students of Yale EEB354 in the fall of 2020. This book started as a collection of lecture notes for this course, the first that I taught fully online. The students provided invaluable motivation, feedback, and patience. Thanks in particular to Lauren Mellenthin, the graduate teaching fellow for the course, for her incredible contributions to this course and, as a result, this book. Members of my lab provided very helpful feedback when I posted new chapters. Steve Haddock and Felipe Zapata provided close reads of host chapters, often within hours of completing first drafts.

## Dedication

This book is dedicated to my sister Jenny Dunn, who has been there every step of the way. I am so grateful for her adventurous spirit, quick mind, and shared love of life. Thank you Jenny.


## Source code

The source code for this manuscript is available on [github]( https://github.com/caseywdunn/phylogenetic_biology). The book was rendered from the source with `bookdown` [@bookdown2016].

Please submit any errors you find, typos, or suggestions that you have for improving the manuscript to the [issue tracker](https://github.com/caseywdunn/phylogenetic_biology/issues).

## Other resources

The following sites have a wide variety of material that is relevant to the 
theory and and practice of phylogenetic biology.

- An extensive list of tools, tutorials, and examples of phylogenetic tools in 
the programming language R maintained by Briam O'Meara:
[CRAN Task View: Phylogenetics](https://cran.r-project.org/web/views/Phylogenetics.html)

- The Workshop on Molecular Evolution at Woods Hole. This is an intensive summer 
course on phylogenetics, with an emphasis on building phylogenetic trees. Check 
out the faculty pages for lecture pdfs:
[Workshop on Molecular Evolution](https://molevolworkshop.github.io)

- The Applied Phylogenetics Workshop in Bodega Bay. This is another summer course 
on phylogenetics, but with a bit more emphasis on using phylogenies to test 
evolutionary questions:
[Applied Phylogenetics Workshop](http://treethinkers.org/tutorials/)

The following books are great resources for learning more about phylogenetic biology:

- Baum, D. and Smith, S. (2012). Tree Thinking. An Introduction to Phylogenetic 
Biology. [Roberts Publishers](http://www.roberts-publishers.com/tree-thinking-an-introduction-to-phylogenetic-biology.html).

- Paradis, E. (2011) Analysis of Phylogenetics and Evolution with R. 
[Springer](http://www.springer.com/life+sciences/evolutionary+%26+developmental+biology/book/978-0-387-32914-7)

- Felsenstein, J. (2004) Inferring phylogenies. 
[Sinauer Associates](http://www.sinauer.com/detail.php?id=1775).

- Swofford, D. L., Olsen, G. J., Waddell, P. J., & Hillis, D. M. (1996). 
Phylogenetic inference. In: Molecular Systematics, Second Edition. eds: D. M. 
Hillis, C Moritz, & B. K. Mable. [Sinauer Associates](http://www.sinauer.com/detail.php?id=1775)

The following books provide general computational background for the topics covered here:

- Wickham, H., Grolemund, G (2017) R for Data Science. https://r4ds.had.co.nz

- Haddock, S. H. D. and Dunn, C. W. (2010). Practical Computing for Biologists. http://practicalcomputing.org


## Software versions

This book was rendered from the source code on $`r format( Sys.time(), "%a %b %d %X %Y" )`$ with the following R package versions.

```{r session_summary, echo=FALSE, comment=NA}
	sessionInfo()
```


## License

This manuscript is distrubuted under a [Creative Commons Attribution-NonCommercial-NoDerivs 3.0 License]( https://creativecommons.org/licenses/by-nc-nd/3.0/us/).
