--- 
title: "Phylogenetic Biology"
author: "Casey W. Dunn"
github-repo: caseywdunn/phylogenetic_biology
twitter-handle: caseywdunn
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "An introduction to Phylogenetic Biology."
fig_caption: yes
---

```{r preliminaries, include=FALSE}
	# The following should be installed from github with the specified 
	# devtools command. You will need to install devtools first.
	library( treeio ) # devtools::install_github("GuangchuangYu/treeio")
	library( ggtree ) # devtools::install_github("GuangchuangYu/ggtree")
	library( ape )
  library( geiger )
	library( gridExtra )
	library( magrittr )
	library( phytools )
	library( stringr )
	library( tidyverse )
	source( "functions.R" )
	set.seed( 23456 )
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE,
  dpi = 300,
  cache = FALSE
  )
```


# Preface {-}

I developed this book as a collection of the concepts, methods, and applications that I most wanted to share with students in my Phylogenetic Biology courses.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'ape', 'geiger', 'ggtree', 'treeio', 'phytools'
), 'packages.bib')
```

## Associated skills

Comfort at the command line, manipulation of text files. *Practical Computing for Biologists*.

Basic literacy with the programming language R. *R for Data Science*.

Basic statistics, in particular probability theory. Some math, in particular an understanding of matrices and manipulations of matrices.




